---
title: "Introduction to Phylogenetic Likelihood"
author: "Bret Larget"
date: "9/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
library(tidyverse)
library(ape)
```

## FASTA format

The FASTA format is a common way to share sequence data.
Each sequence begins with a line that starts with the key character `>` followed by the name of the species and/or other identifying information
and then a line break.
Subsequent lines have the sequence data which may be broken across multiple lines.
The next sequence begins on another row beginning with `>`.

## The ape package

The ape package has a number of helpful phylogenetic functions, but we will just use `ape::read.dna()` to read in the FASTA sequence data.
By default, ape converts DNA into a binary representation.
It will be easier for us to have the character data.

```{r}
sg = read.dna("sheep-giraffe.fasta", as.character=TRUE, format="fasta")
sg[,1:20]
table(sg[1,],sg[2,])
```

## Q Matrix

This code produces a crude estimate of a Q-matrix from the data
which can be used for the likelihood estimate. 

```{r q-matrix, include=FALSE}
p = table(sg) / sum(table(sg));p
tab = table(sg[1,],sg[2,])
n = sum(tab)
rho = (tab + t(tab))[row(tab) > col(tab)]
rho = rho/sum(rho)
Q = matrix(0,4,4)
Q[row(Q) > col(Q)] = rho
Q = Q + t(Q)
Q = diag(1/p) %*% Q
Q = Q/2
diag(Q) = -apply(Q,1,sum)
Q
```

## Probability Transition Matrix

For a given nonnegative value $t$,
define
$$
P(t) = e^{Qt}
$$

If $Q$ has eigendecomposition $Q = V\Lambda V^{-1}$,
then
$$
P(t) = V e^{\Lambda t} V^{-1}
$$

## Likelihood

The likelihood of the time $t$ given the data and the $Q$ matrix is
$$
L = \prod_{i=1}^n \pi_{x[i]} P_{x[i],y[i]}(0.1)
$$
where $\pi$ is the stationary distribution of $Q$
and $x$ and $y$ are the two sequences.

As an example, the likelihood of the first site (AA) over a branch of length $t=0.1$ can be calculated like this.

```{r like1}
e = eigen(Q)
lambda = e$values
V = e$vectors
Vinv = solve(V)
P = V %*% diag(exp(lambda*0.1)) %*% Vinv
tol = 1e-12
m = which(abs(lambda) < tol)
pii = Vinv[m,] / sum(Vinv[m,])
like1 = pii[1] * P[1,1]
like1
```

The log-likelihood is the sum of the logs of the likelihoods calculated over each site (column of the sequence data).

## Exercises

1. Write code to find the log-likelihood of the data using the given $Q$ and $\pi$ as a function of $t$ and plot this from $t=0.01$ to $t=0.5$.

```{r expand-Q}
expand_Q = function(Q)
{
  e = eigen(Q)
  lambda = e$values
  V = e$vectors
  Vinv = solve(V)
  tol = 1e-12
  m = which(abs(lambda) < tol)
  p = Vinv[m,] / sum(Vinv[m,])
  out = list(Q=Q,V=V,Vinv=Vinv,lambda=lambda,pi=p)
  return(out)
}
```

```{r table-counts}
## only works when x and y are vectors of a, c, g, and t
## of the same length
get_counts = function(x,y)
{
  return( table(x,y))
}
```

```{r log-like}
## qlist is the eigendecomposition and related parts of Q
qlist = expand_Q(Q)
## counts for sheep and giraffe
n = get_counts(sg[1,],sg[2,])

## t0 is the branch length
logl = function(t0,qlist,n)
{
  P = qlist$V %*% diag(exp(qlist$lambda*t0)) %*% qlist$Vinv
  seq1_total = apply(n,1,sum)
  sum1 = sum(seq1_total * log(qlist$pi))
  sum2 = sum(n * log(P))
  return( sum1 + sum2 )
}
```

```{r example}
t0 = seq(0.1,0.2, length.out = 101)
y = numeric(101)
for ( i in 1:101 )
{
  y[i] = logl(t0[i],qlist,n)
}

x = tibble(t=t0,logl=y)

ggplot(x, aes(x=t0,y=y)) +
  geom_line()
```

2. Write code to find the maximum likelihood estimate of $t$.

```{r likelihood}
## likelihood at t = 0.01
P1 =  V %*% diag(exp(lambda*0.01)) %*% Vinv
P = V %*% diag(exp(lambda*0.03)) %*% Vinv
P2 = P1 %*% P1
P3 = P2 %*% P1
tab_rowsum = apply(tab,1,sum)
log_like = sum(log(pii) * tab_rowsum) + sum(tab * log(P1))

```


### Optimize

```{r optimize}
opt1 = optimize(interval=c(0,1),maximum=TRUE,logl,qlist,n)
```

### Estimate all Parameters

Parameters are:

- $\pi = (\pi_A,\pi_C,\pi_G,\pi_T)$ where $\sum_i \pi_i=1$
- $\rho = (\rho_{AC},\ldots,\rho_{GT})$ where $\sum_i \sum_{j \neq i} \rho_{ij} = 1$ and $\rho_{ij} = \rho_{ji}$

$Q$-Matrix:

- $Q_{ij} =$
    - $\rho_{ij}/(2\pi_i)$ if $i \neq j$
    - $\rho_{i+}/(2\pi_i)$ if $i=j$
    
where $\rho_{i+} = \sum_{j \neq i} \rho_{ij}$   

### Parameterize

Put all free parameters in one vector.

$$
\theta = (t,\pi_A,\pi_C,\pi_G,\rho_{AC},\rho_{AG},\rho_{AT},\rho_{CG}, \rho_{CT})
$$

and let 

- $\pi_T = 1 - (\pi_A + \pi_C + \pi_G)$
- $\rho_{GT} = 1 - (\rho_{AC} +\rho_{AG} + \rho_{AT} + \rho_{CG} + \rho_{CT})$

### New Log-Likelihood Function

```{r full-log-likelihood}
## code to initialize from data
get_n = function(sequences)
{
  n = table(sequences[1,],sequences[2,])
  return( n )
}

get_pi = function(sequences)
{
  p = table(sequences)
  return( p/sum(p) )
}

get_rho = function(n)
{
  rho = (n + t(n))[row(n) > col(n)]
  names(rho) = c("ac","ag","at","cg","ct","gt")
  return( rho / sum(rho) )
}

get_Qlist = function(pi,rho)
{
  Q = matrix(0,4,4)
  Q[row(Q) > col(Q)] = rho
  Q = Q + t(Q)
  Q = diag(1/pi) %*% Q/2
  diag(Q) = -apply(Q,1,sum)
  e = eigen(Q)
  lambda = e$values
  V = e$vectors
  Vinv = solve(V)
  return(list(Q=Q,pi=pi,rho=rho,V=V,Vinv=Vinv,lambda=lambda))
}

## theta = c(t,z1,z2,z3,w1,w2,w3,w4,w5)
## where:
##   t is the branch length
##   z1 = pi[1]
##   z2 = pi[2] / (1 - pi[1])
##   z3 = pi[3] / (1 - pi[1] - pi[2])
##   w1 = rho[1]
##   ... and so on

simplex_to_box = function(x)
{
  k = length(x)
  z = numeric(k-1)
  s = cumsum(x)
  z = x[1:(k-1)] / (1 - c(0,s[1:(k-2)]))
  return(z)
}

box_to_simplex(z)
{
  x = ...
  return(x)
}

get_initial_theta = function(sequences)
{
  ## observed matrix of site pattern counts
  n = get_n(sequences)
  ## observed base relative frequencies
  p = get_pi(sequences)
  ## observed relative frequencies of changed changed sites
  rho = get_rho(n)
  ## JC1969 distance
  p_diff = (sum(n) - sum(diag(n))) / (sum(n))
  t0 = -0.75 * log(1 - 4*p_diff/3)
  theta = c(t0,p[1:3],rho[1:5])
  return( theta )
}

get_parameters = function(theta)
{
  t0 = theta[1]
  p = c(theta[2:4],1-sum(theta[2:4]))
  rho = c(theta[5:9],1-sum(theta[5:9]))
  ## estimated Q matrix and decomposition
  qlist = get_Qlist(p,rho)
  
  return( list(t=t0,pi=p,rho=rho,lambda=qlist$lambda,V=qlist$V,Vinv=qlist$Vinv))
}

## theta a vector as above where t>0, all other parameters between 0 and 1.
## n is the matrix of counts
logl_full = function(theta,n)
{
  ## p_list has $t, $pi, $rho, $V, $Vinv, $lambda
  p_list = get_parameters(theta)
  logP = log(p_list$V %*% diag(exp(p_list$t * p_list$lambda)) %*% p_list$Vinv)
  seq1_total = apply(n,1,sum)
  sum1 = sum(seq1_total * log(p_list$pi))
  sum2 = sum(n * logP)
  return( sum1 + sum2 )
}
```

## Optimize

```{r optimize-gtr}
out = optim(get_initial_theta(sg),logl_full,method="L-BFGS-B",
            lower = rep(0,9),
            upper = c(5,rep(1,8)),
            control = list(fnscale=-1),
            n=n)
```
